---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![Integrative Methods of Analysis for Genetic Epidemiology](https://raw.githubusercontent.com/USCbiostats/badges/master/tommy-image-badge.svg)](https://image.usc.edu)

# geese: *GE*ne-functional *E*volution using *S*uffici*E*ncy <img src="man/figures/logo.svg" align="right" width="180px"/>

<!-- badges: start -->
<!-- badges: end -->

The goal of geese is to ...

## Installation

<!-- You can install the released version of geese from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("geese") -->
<!-- ``` -->

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("USCbiostats/geese")
```
# Examples

## Simulating annotations (two different sets)

```{r simulating-data}
library(geese)

# Preparing data
n <- 100L
annotations <- replicate(n * 2 - 1, c(9, 9), simplify = FALSE)

# Random tree
set.seed(31)
tree <- aphylo::sim_tree(n)$edge - 1L

duplication <- rep(TRUE, n * 2 - 1)

# Reading the data in
amodel <- new_geese(
  annotations = annotations,
  geneid = c(tree[, 2], n),
  parent = c(tree[, 1], -1),
  duplication = duplication
  )

# Preparing the model
term_gains(amodel, 0:1)
term_loss(amodel, 0:1)
term_maxfuns(amodel, 1, 1)
init_model(amodel)

# Testing
params <- c(
  # Gains
  2, 1.5,
  # Loss
  -2, -1.5,
  # Max funs
  2, 
  # Root probabilities
  -10, -10
)
names(params) <- c("gain0", "gain1", "loss0", "loss1", "onefun", "root0", "root1")

likelihood(amodel, params*0) # Equals 1 b/c all missings

# Simulating data
fake1 <- sim_geese(p = amodel, par = params, seed = 1110)
fake2 <- sim_geese(p = amodel, par = params)
```

We can now visualize either of the annotations using the [aphylo](https://github.com/USCbiostats/aphylo) package.

```{r viz-with-aphylo}
library(aphylo)
ap <- new_aphylo(
  tree           = {set.seed(31);aphylo::sim_tree(n)},
  tip.annotation = as.data.frame(do.call(rbind, fake1[1:n]))
)
plot(ap)
```

## Model fitting MLE

```{r mle}
# Creating the object
amodel <- new_geese(
  annotations = fake1[c(tree[, 2], n) + 1],
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication
  )

# Adding the model terms
term_gains(amodel, 0:1)
term_loss(amodel, 0:1)
term_maxfuns(amodel, 1, 1)

# We need to initialize to do all the accountintg
init_model(amodel)

# Finding MLE
geese_mle(amodel, hessian = TRUE)
```

## Model fitting MCMC

```{r mcmc-fit, cache=TRUE}
set.seed(122)
ans_mcmc <- geese_mcmc(
  amodel,
  nsteps  = 40000,
  kernel  = fmcmc::kernel_ram(warmup = 2000), 
  prior   = function(p) dlogis(p, scale = 2, log = TRUE)
  )
```

We can take a look at the results like this:

```{r mcmc-analysis, echo = FALSE}
op <- par(
  mfrow = c(4, 2), #tcl=.5,
  las=1, mar = c(3,3,1,0),
  bty = "n", oma = rep(1,4)
  )
for (i in 1:ncol(ans_mcmc)) {
  tmpx <- window(ans_mcmc, start = 20000)[,i,drop=FALSE]
  
  coda::traceplot(
    tmpx, smooth = FALSE, ylim = c(-11,11), col = rgb(0, 128, 128, maxColorValue = 255), 
    main = names(params)[i]
    )
  abline(h = params[i], lty=3, lwd=4, col = "red")
}
par(op)
summary(window(ans_mcmc, start = 15000))
```

## Using a flock

GEESE models can be grouped (pooled) into a flock.

```{r new-flock}
flock <- new_flock()

# Adding first set of annotations
add_geese(
  flock,
  annotations = fake1[c(tree[, 2], n) + 1],
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication  
)

# Now the second set
add_geese(
  flock,
  annotations = fake2[c(tree[, 2], n) + 1],
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication  
)

# Adding the model terms
term_gains(flock, 0:1)
term_loss(flock, 0:1)
term_maxfuns(flock, 1, 1)

# We need to initialize to do all the accountintg
init_model(flock)
```

We can use the same program to fit the MCMC

```{r flock-mcmc, cache=TRUE}
set.seed(122)
ans_mcmc2 <- geese_mcmc(
  flock,
  nsteps  = 20000,
  kernel  = fmcmc::kernel_ram(warmup = 2000), 
  prior   = function(p) dlogis(p, scale = 2, log = TRUE)
  )
```

```{r viz-flock}
op <- par(
  mfrow = c(4, 2), #tcl=.5,
  las=1, mar = c(3,3,1,0),
  bty = "n", oma = rep(1,4)
  )
for (i in 1:ncol(ans_mcmc2)) {
  tmpx <- window(ans_mcmc2, start = 10000)[,i,drop=FALSE]
  
  coda::traceplot(
    tmpx, smooth = FALSE, ylim = c(-11,11), col = rgb(0, 128, 128, maxColorValue = 255), 
    main = names(params)[i]
    )
  abline(h = params[i], lty=3, lwd=4, col = "red")
}
par(op)
summary(window(ans_mcmc2, start = 10000))
```


## Code of Conduct
  
Please note that the aphylo2 project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
