---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

[![Integrative Methods of Analysis for Genetic Epidemiology](https://raw.githubusercontent.com/USCbiostats/badges/master/tommy-image-badge.svg)](https://image.usc.edu)

# geese: *GE*ne-functional *E*volution using *S*uffici*E*ncy <img src="man/figures/logo.svg" align="right" width="180px"/>

<!-- badges: start -->
[![R-CMD-check](https://github.com/USCbiostats/geese/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/USCbiostats/geese/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This R package taps into statistical theory mostly developed in the social networks
realm. Using Exponential-Family Random Graph Models (ERGMs), `geese` provides an
statistical framework for building Gene Functional Evolution Models using Sufficiency.
For example, users can directly hypothesize whether Neofunctionalization or Subfunctionalization
events were taking place in a phylogeny, without having to estimate the full transition
Markov Matrix that is usually used.

GEESE is computationally efficient, with C++ under-the-hood, allowing the analyses
of either single trees (a GEESE) or multiple trees simulatenously (pooled model),
in a Flock.

This is work in progress and based on the theoretical work developed duing
[George G. Vega Yon](https://ggv.cl)'s doctoral thesis.

## Installation

<!-- You can install the released version of geese from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("geese") -->
<!-- ``` -->

The development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("USCbiostats/geese")
```
# Examples

## Simulating annotations (two different sets)

```{r simulating-data}
library(geese)

# Preparing data
n <- 100L
annotations <- replicate(n * 2 - 1, c(9, 9), simplify = FALSE)

# Random tree
set.seed(3151)
tree <- aphylo::sim_tree(n)$edge - 1L

duplication <- c(rep(FALSE, n), sample.int(2, n - 1, replace = TRUE, prob = c(.8, .2)) == 2)

# Reading the data in
amodel <- new_geese(
  annotations = annotations,
  geneid = c(tree[, 2], n),
  parent = c(tree[, 1], -1),
  duplication = duplication
  )

# Persistence to preserve parent state
term_overall_changes(amodel, duplication = 0)
term_overall_changes(amodel, duplication = 1)

# Tendency for only one sib gaining function 0
term_pairwise_neofun_singlefun(amodel, 0, duplication = 1)

# Function 1 is gain from function 0
term_neofun_a2b(amodel, 0, 1, duplication = 1)

init_model(amodel)

# Testing
params <- c(
  # Persist the parent state
  -4, -4, 
  # Gain function 0
  3, 
  # Neofun 0 -> 1
  5,
  # Root probabilities
  -10, -10
)
names(params) <- c(names(amodel), "root0", "root1")

likelihood(amodel, params*0) # Equals 1 b/c all missings

# Simulating data
fake1 <- sim_geese(p = amodel, par = params, seed = 1110)
fake2 <- sim_geese(p = amodel, par = params)
```

We can now visualize either of the annotations using the [aphylo](https://github.com/USCbiostats/aphylo) package.

```{r viz-with-aphylo}
library(aphylo)
ap <- aphylo_from_data_frame(
  tree        = as.phylo(tree), 
  annotations = data.frame(
    id = c(tree[, 2], n),
    do.call(rbind, fake1)
    ),
  types = data.frame(0:(2*n-2), as.integer(!duplication))
)
plot(ap)
```

## Model fitting MLE

```{r mle}
# Creating the object
amodel <- new_geese(
  annotations = fake1,
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication
  )

# Persistence to preserve parent state
term_overall_changes(amodel, duplication = 0)
term_overall_changes(amodel, duplication = 1)

# Tendency for only one sib gaining function 0
term_pairwise_neofun_singlefun(amodel, 0, duplication = 1)

# Function 1 is gain from function 0
term_neofun_a2b(amodel, 0, 1, duplication = 1)

# We need to initialize to do all the accounting
init_model(amodel)

print(amodel)

# Finding MLE
geese_mle(amodel, hessian = TRUE)

transition_prob(
  amodel,
  params = params[-((length(params) - 1):length(params))], 
  duplication = TRUE, state = c(FALSE, FALSE),
  array = matrix(c(1, 0, 0, 1)*0, ncol=2)
)
```

## Model fitting MCMC

```{r mcmc-fit, cache=TRUE}
set.seed(122)
ans_mcmc <- geese_mcmc(
  amodel,
  nsteps  = 20000,
  kernel  = fmcmc::kernel_ram(warmup = 2000), 
  prior   = function(p) dnorm(p, sd = 2, log = TRUE)
  )
```

We can take a look at the results like this:

```{r mcmc-analysis, echo = FALSE}
mcmc_estimates <- ans_mcmc # oldans$estimates
op <- par(
  mfrow = c(4, 2), #tcl=.5,
  las=1, mar = c(3,3,1,0),
  bty = "n", oma = rep(1,4)
  )
for (i in 1:ncol(mcmc_estimates)) {
  tmpx <- window(mcmc_estimates, start = 10000)[,i,drop=FALSE]
  
  coda::traceplot(
    tmpx, smooth = FALSE, ylim = c(-11,11), col = rgb(0, 128, 128, maxColorValue = 255), 
    main = names(params)[i]
    )
  abline(h = params[i], lty=3, lwd=4, col = "red")
}
par(op)
summary(window(mcmc_estimates, start = end(ans_mcmc) * 3/4))
```

```{r prediction}
# Obtaining estimates
par_estimates <- colMeans(window(ans_mcmc, start = end(ans_mcmc) * 3/4))
ans_pred      <- predict_geese(amodel, par_estimates, leave_one_out = TRUE)
ans_pred      <- do.call(rbind, ans_pred)

# Preparing annotations
ann_obs <- do.call(rbind, fake1)

# Subsetting
idx <- which(tree[, 2] < 100)
ans_pred <- ans_pred[idx,]
ann_obs <- ann_obs[idx,]

# Mean Absolute Error
hist(abs(ans_pred - ann_obs))

# AUC
(ans <- aphylo::prediction_score(
  cbind(as.vector(ans_pred)),
  cbind(as.vector(ann_obs))
  ))
plot(ans$auc)
```


<!-- We can compare these results to what we would obtain using aphylo -->

<!-- # ```{r aphylo} -->
<!-- # ans_aphylo <- aphylo_mcmc(ap ~ mu_d + mu_s + Pi) -->
<!-- # (ps <- prediction_score(ans_aphylo)) -->
<!-- # ``` -->


## Using a flock

GEESE models can be grouped (pooled) into a flock.

```{r new-flock}
flock <- new_flock()

# Adding first set of annotations
add_geese(
  flock,
  annotations = fake1,
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication  
)

# Now the second set
add_geese(
  flock,
  annotations = fake2,
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication  
)

# Persistence to preserve parent state
term_overall_changes(amodel, duplication = 0)
term_overall_changes(amodel, duplication = 1)

# Tendency for only one sib gaining function 0
term_pairwise_neofun_singlefun(amodel, 0, duplication = 1)

# Function 1 is gain from function 0
term_neofun_a2b(amodel, 0, 1, duplication = 1)

# We need to initialize to do all the accountintg
init_model(flock)

print(flock)
```

We can use the same program to fit the MCMC

```{r flock-mcmc, cache=TRUE}
set.seed(122)
ans_mcmc2 <- geese_mcmc(
  flock,
  nsteps  = 20000,
  kernel  = fmcmc::kernel_ram(warmup = 2000), 
  prior   = function(p) dlogis(p, scale = 2, log = TRUE)
  )
```

```{r viz-flock}
op <- par(
  mfrow = c(4, 2), #tcl=.5,
  las=1, mar = c(3,3,1,0),
  bty = "n", oma = rep(1,4)
  )
for (i in 1:ncol(ans_mcmc2)) {
  tmpx <- window(ans_mcmc2, start = 10000)[,i,drop=FALSE]
  
  coda::traceplot(
    tmpx, smooth = FALSE, ylim = c(-11,11), col = rgb(0, 128, 128, maxColorValue = 255), 
    main = names(params)[i]
    )
  abline(h = params[i], lty=3, lwd=4, col = "red")
}
par(op)
summary(window(ans_mcmc2, start = 10000))
```

## Limiting the support

In this example, we use the function `rule_limit_changes()` to apply a
constraint to the support of the model. This takes the first 2 term (0 and 1 since the
index is in C++), and restricts the support to states where there are between
$[0, 2]$ changes, at most.

This should be useful when dealing with multiple functions or [pylotomies](https://en.wikipedia.org/wiki/Polytomy).

```{r limit_support}
# Creating the object
amodel_limited <- new_geese(
  annotations = fake1,
  geneid      = c(tree[, 2], n),
  parent      = c(tree[, 1],-1),
  duplication = duplication
  )

# Persistence to preserve parent state
term_overall_changes(amodel_limited, duplication = 0)
term_overall_changes(amodel_limited, duplication = 1)

# Tendency for only one sib gaining function 0
term_pairwise_neofun_singlefun(amodel_limited, 0, duplication = 1)

# Function 1 is gain from function 0
term_neofun_a2b(amodel_limited, 0, 1, duplication = 1)

# At most two moves
rule_limit_changes(amodel_limited, 0, 0, 2, duplication = 0)
rule_limit_changes(amodel_limited, 1, 0, 2, duplication = 1)

# We need to initialize to do all the accounting
init_model(amodel_limited)

# Is limiting the support any useful?
support_size(amodel_limited)
```

Since we added the constraint based on the term `term_overall_changes()`, we
now need to fix the parameter at 0 (i.e. no effect) during the MCMC model:

```{r limit-mcmc, cache = TRUE}
set.seed(122)
ans_mcmc2 <- geese_mcmc(
  amodel_limited,
  nsteps  = 20000,
  kernel  = fmcmc::kernel_ram(
    warmup = 2000 #,
    # fixed  = c(TRUE, TRUE, rep(FALSE, nterms(amodel_limited) - 2))
    ), 
  prior   = function(p) dnorm(p, sd = 2, log = TRUE)
  )
```

```{r mcmc-analysis-limited, echo = FALSE}
op <- par(
  mfrow = c(4, 2), #tcl=.5,
  las=1, mar = c(3,3,1,0),
  bty = "n", oma = rep(1,4)
  )
for (i in 1:ncol(ans_mcmc2)) {
  tmpx <- window(ans_mcmc2, start = 10000)[,i,drop=FALSE]
  
  coda::traceplot(
    tmpx, smooth = FALSE, ylim = c(-11,11), col = rgb(0, 128, 128, maxColorValue = 255), 
    main = names(params)[i]
    )
  abline(h = params[i], lty=3, lwd=4, col = "red")
}
par(op)
summary(window(ans_mcmc2, start = 15000))
```

# Code of Conduct
  
Please note that the aphylo2 project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
